---
title: "Project Pt2 Price"
author: "Group 1"
date: "2026-03-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pre-Project Part 2 filtering and processing
Delivery Data

```{r}
library(readr)
library(ggplot2)
library(readr)
library(ggplot2)
library(quantmod) # loading data sets from online
library(dplyr) # for the pipes
library(kableExtra) # for kable
library(lubridate) # for years...
library(tidyr)
library(knitr)
library(forecast)     # for Acf, Pacf, auto.arima, etc.  
library(tseries)      # for adf.test  
library(gridExtra)    # for arranging plots  


data2 <- read_csv("3210035101_databaseLoadingData_2.csv")
head(data2)



delivery_data = data.frame(Date = data2$REF_DATE,
                           Value = as.numeric(data2$VALUE))

head(delivery_data)
p2 = ggplot(delivery_data, aes(x=Date, y = Value, group=1))+
             geom_line(color="blue", size=1)+
  labs(
    title = "Producer deliveries of Wheat, Excluding Durum",
    subtitle = "2002-2025",
    x = "Date (Monthly)",
    y = "Grain (Tonnes)",
    caption = "Source: Statistics Canada. Table 32-10-0351-01  Producer deliveries of major grains"
  ) 
print(p2)

delivery_ts <- ts(data2$VALUE,
                  start = c(2002,1), # Start year and month: January 2002
                  frequency = 12) # 12 observations per year (monthly data)


# Define moving average filter function
moving_average <- function(X, q) {
  N <- length(X)              # Length of input series
  weight <- 1 / (2 * q + 1)   # Equal weights for symmetric MA
  indices <- (q+1):(N-q)      # Indices where filter can be applied
  X_filt <- rep(NA, N)        # Initialize output with NAs
  
  # Apply moving average filter
  for (i in indices) {
    X_filt[i] <- sum(X[(i-q):(i+q)] * weight)
  }
  return(X_filt)               # Return filtered series
}


delivery_data_D3 = diff(log(delivery_data$Value))

delivery_data_des = diff(delivery_data_D3, lag=12)

```

Price Data
```{r}
library(patchwork)
library(tseries)
library(gridExtra)
library(readr)
library(ggplot2)
library(quantmod) # loading data sets from online
library(dplyr) # for the pipes
library(kableExtra) # for kable
library(lubridate) # for years...
library(knitr)
library(tidyr)

Data_Price <- read.csv("https://raw.githubusercontent.com/stardustsunrise/Time_Series/refs/heads/main/3210007701_databaseLoadingData.csv")

head(Data_Price)
str(Data_Price)

print(Data_Price)


#Convert to data frame


Price_data = data.frame(Date= as.Date(paste0(Data_Price$REF_DATE,"-01")), 
                        Price_per_MTonne= Data_Price$VALUE,
                        
                        Product=c("Wheat (except durum wheat) [1121111]"))
print(Price_data)

# Averaging the price of "Wheat (except durum wheat) [1121111]" accross all provinces to have 1 avg monthly price value for whole of Canada
Price_data<-Price_data%>%
  group_by(Date)%>%
  summarise(avg_price_canada=mean(Price_per_MTonne,na.rm= TRUE))%>%
  mutate(Months=row_number(),
         Year=year(Date),
         Quarter=quarter(Date))%>%
  ungroup()

print(Price_data)
        
#2.1 Removing Trend

#converting time series to object

Price_data_ts=ts(Price_data$avg_price_canada,
                 start = c(2002-01-01,1),
                 frequency = 12)

str(Price_data_ts)

#calculate transformation
Diff=diff(Price_data_ts)#First difference
Log_diff=diff(log(Price_data_ts))#Difference of Logarithms
#2.2 Removing seasonal component

library(forecast)


# Removing Seasonality

P_D1<-diff(Log_diff,difference=1)
P_D3<-diff(Log_diff,differences=3)
P_D12<-diff(Log_diff,differences=12)
P_D24<-diff(Log_diff,differences=24)


#preference to Difference of 12th order

Deseasonlised_pricedata=P_D12

# Apply filtering to log-differenced series
Series_to_filter <- Price_data_ts
n <- length(Series_to_filter)

# --- q = 6 ---
q1 <- 6 #this is in the range from 5 to 9, helping reduce the high volatility of the data
weights1 <- rep(1/(2*q1+1), 2*q1+1)

Linear_filtered_q6 <- stats::filter(
  Series_to_filter,
  filter = weights1,
  sides = 2
)

# --- q = 12 ---
q2 <- 12 #this helps to address the seasonality of the data
weights2 <- rep(1/(2*q2+1), 2*q2+1)

Linear_filtered_q12 <- stats::filter(
  Series_to_filter,
  filter = weights2,
  sides = 2
)
```
```{r}
#compare the differences
plot(P_D1)
plot(P_D12)
```
```{r}
library(forecast)
library(gridExtra)
library(ggplot2)

# Function to simulate and return ACF/PACF plots
plot_arma_acf_pacf <- function(ar = numeric(0), ma = numeric(0), n = 500) {
  # Use safe coefficients if ar/ma supplied are too big
  if(length(ar) > 0) ar <- rep(0.3, length(ar))
  if(length(ma) > 0) ma <- rep(0.5, length(ma))
  
  ts_data <- arima.sim(model = list(ar = ar, ma = ma), n = n)
  acf_plot <- ggAcf(ts_data) + ggtitle(paste0("ACF AR=", length(ar), " MA=", length(ma)))
  pacf_plot <- ggPacf(ts_data) + ggtitle(paste0("PACF AR=", length(ar), " MA=", length(ma)))
  return(list(acf_plot, pacf_plot))
}

# Example: simulate ARMA(2,1)
plots <- plot_arma_acf_pacf(ar = c(0.3, 0.3), ma = 0.5)
grid.arrange(plots[[1]], plots[[2]], ncol = 2)
```


# Question 5: Dickey-Fuller Unit Root Test
Delivery Data
```{r}

library(tseries)

# ADF Test
adf.test(delivery_data_des)

```

Price Data
```{r}

Stationary_series <- Deseasonlised_pricedata
library(tseries)
# ADF Test
adf.test(Stationary_series)
```
p-value = 0.01 < 0.05 and thus we confirm that the data is stationary with 95% certainty
# Question 6: Selection of ARMA Model Orders

Delivery Data

```{r}
#---ACF estimator function---------------

ACF_estimator <- function(X, hmax){
      n <- length(X)
      X_bar <- mean(X)
      acf_vals <- numeric(hmax+1)
      
      for(h in 0:hmax){
        num <- sum((X[(h+1):n]-X_bar)*
                     (X[1:(n-h)]-X_bar))
        den <- sum((X-X_bar)^2)
        acf_vals[h+1] <- num/den
      }
      
      return(acf_vals)
    }


res <- na.omit(delivery_data_des)

hmax <- 40


acf_values <- ACF_estimator(res, hmax)

#---PACF estimator function-----------------------------------

PACF_estimator <- function(X, hmax){
      
      X <- na.omit(X)        # ← critical line
      pacf_vals <- numeric(hmax)
      
      for(h in 1:hmax){
        model <- ar(X, order.max = h, aic = FALSE, method = "yw")
        pacf_vals[h] <- model$ar[h]
      }
      
      return(pacf_vals)
    }


pacf_values <- PACF_estimator(delivery_data_des, hmax)

#--------Setting Up the CI--------------

CI_95 <- 1.96 / sqrt(length(res))  #90% 1.645    99% 2.576
CI_90 <- 1.645 / sqrt(length(res))
CI_99 <- 2.576 / sqrt(length(res))

#----------Plot ACF-----------------------------------

#90% CI Plot

plot(0:hmax, acf_values, type = "h",
      main = "Estimated ACF 90% CI",
      xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_90, col = "red", lty = 2)
abline(h = -CI_90, col = "red", lty = 2)


    #95% CI Plot
plot(0:hmax, acf_values, type = "h",
     main = "Estimated ACF 95% CI",
     xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_95, col = "red", lty = 2)
abline(h = -CI_95, col = "red", lty = 2)

#99% CI Plot
plot(0:hmax, acf_values, type = "h",
      main = "Estimated ACF 99% CI",
      xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_99, col = "red", lty = 2)
abline(h = -CI_99, col = "red", lty = 2)

#----------Plot PACF-------------------------

#90% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 90% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_90, col="blue", lty=2)
abline(h=-CI_90, col="blue", lty=2)


#95% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 95% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_95, col="blue", lty=2)
abline(h=-CI_95, col="blue", lty=2)


#99% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 99% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_99, col="blue", lty=2)
abline(h=-CI_99, col="blue", lty=2)




```
Point 3

The ACF shows a significant spike at Lag 1 and a potentially significant spike at Lag 2 before cutting off. This is characteristic of a Moving Average process, specifically suggesting an MA(1) or MA(2). The PACF does not show a clear cutoff; instead, we can see a decaying pattern (ignoring seasonal lags at 12, 24, and 36). This decay in the PACF is a sign of an AR(1) component or that this is simply and MA(q) process.

Significant spikes remain at seasonal lags (12, 24, 36). While these indicate remaining seasonal correlation, they are being bypassed for the scope of this project, as advised by professor Boire.

For Question 7, we compare three model ARMA(1,2), ARMA(1,1) and MA(2).


Price Data

##(1) - ACF and PACF and (2) - CIs
Autocorrelation Function (ACF):
$$ρ^​(h)=∑t=1n​(Xt​−Xˉ)2∑t=h+1n​(Xt​−Xˉ)(Xt−h​−Xˉ)​$$
```{r}
ACF_estimator <- function(X, hmax){
  n <- length(X)
  X_bar <- mean(X)
  acf_vals <- numeric(hmax+1)
  
  for(h in 0:hmax){
    num <- sum((X[(h+1):n]-X_bar)*
                 (X[1:(n-h)]-X_bar))
    den <- sum((X-X_bar)^2)
    acf_vals[h+1] <- num/den
  }
  
  return(acf_vals)
}

hmax <- 40
acf_values <- ACF_estimator(Stationary_series, hmax)

plot_acf_ci <- function(acf_values, series, hmax, ci = 0.95) {
  
  n <- length(series)
  
  # get correct z value
  z <- qnorm(1 - (1 - ci) / 2)
  
  # confidence interval
  CI <- z / sqrt(n)
  
  plot(0:hmax, acf_values, type = "h",
       main = paste("Estimated ACF (", ci*100, "% CI)", sep=""),
       xlab = "Lag", ylab = "ACF")
  
  abline(h = 0)
  abline(h = CI, col = "red", lty = 2)
  abline(h = -CI, col = "red", lty = 2)
}
# Plots for different CIs
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.90)
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.95)
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.99)

```
Lags for which coefficients not statistically different from 0:
90%: 0,1,2,12,22,24,26,30,36,38; 95%: 0,1,2,12,22,24,30,36,38; 99%: 1,2,12,22,24,36,38.

Partial Autocorelation function (PACF):
```{r}
PACF_estimator <- function(X, hmax){
  pacf_vals <- numeric(hmax)
  
  for(h in 1:hmax){
    model <- ar(X, order.max=h, aic=FALSE, method="yw")
    pacf_vals[h] <- model$ar[h]
  }
  
  return(pacf_vals)
}


# estimator with chosen CIs
plot_pacf_ci <- function(pacf_values, series, hmax, ci = 0.95){
  
  n <- length(series)
  z <- qnorm(1 - (1 - ci)/2)
  CI <- z / sqrt(n)
  
  plot(1:hmax, pacf_values, type="h",
       main=paste("Estimated PACF (", ci*100, "% CI)", sep=""),
       xlab="Lag", ylab="PACF")
  
  abline(h=0)
  abline(h=CI, col="blue", lty=2)
  abline(h=-CI, col="blue", lty=2)
}
pacf_values <- PACF_estimator(Stationary_series, hmax)

plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.90)
plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.95)
plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.99)

```
Lags for which coefficients not statistically different from 0:
90%:1,2,3,4,5,6,9,10,11,12,22,23,25,29,31,35,37; 95%: 1,2,3,4,9,10,11,12,22,35; 99%:1,2,3,4,10,11,35.


##(3) - Hypothesis regarding appropriate orders of the ARMA(p,q)


```{r}

auto.arima(Stationary_series, seasonal = FALSE, stationary =TRUE)
```
The test says that this series matches an AR(5) model. However,the graphs suggest more of a ARMA(3,4) model (4 significant lags on ACF, 3 significant lags on PACF). We will try to go off of the graph for the sake of this analysis.

#Question 7

Delivery Data
```{r stargazer, results='asis'}

#-----Splitting into training test and test sets----------------
n = length(delivery_data_des)

y = delivery_data_des

n_train <- floor(0.8 * n)
y_train <- y[1:n_train]
y_test <- y[(n_train+1):n]


#----Model Estimation on the Training Set--------------------

fit_11 <- arima(y_train, order = c(1,0,2), include.mean = FALSE) #fit given by 
fit_12 <- arima(y_train, order = c(1,0,1), include.mean = FALSE) #testing this fit to see if we're overfitting
fit_13 <- arima(y_train, order = c(0,0,2), include.mean = FALSE) #random test


#----Displaying the results with stargazer---------

se_11 <- sqrt(diag(fit_11$var.coef))
se_12 <- sqrt(diag(fit_12$var.coef))
se_13 <- sqrt(diag(fit_13$var.coef))

library(stargazer)

stargazer(fit_11, fit_12, fit_13,
type = "latex",
se = list(se_11, se_12, se_13),
title = "ARMA Estimation Results",
column.labels = c("ARMA(1,2)", "ARMA(1,1)", "MA(2)"),
digits = 3,
header = FALSE)


```
For the model ARMA(1,2), all of our coefficients are significative. A strong negative autoregressive coefficient AR(1)= -0.886 suggests the series is oscillating. If the value was high last month, it’s very likely to be low this month.MA(1) = 0.304 is a momentum effect and if we had a positive shock last month, approximately 30.4% of that shock persists into the next month in the same direction. The second lag MA(2) = 0.696 is the most dominant feature of the moving average component. A positive shock occurring at two months ago would results in a significant downward pressure (around 69.9% of that positive shock) on the series this month.

For the second model ARMA(1,1), The AR(1)=0.056 is not significant. This model is "over-specified." It is trying to use an autoregressive component that isn't actually helping.

For the third model, the MA(2)= 0.048 term is not significant for similar reasons as the ARMA(1,1)

The best model so far is the ARMA(1,2).


Price Data
Checking for structural breaks
```{r}
plot(Stationary_series, type="l")
```
While there are some spikes in the data, it does not appear that there were any structural breaks. Thus, we can proceed with an 80-20 division for the data
##(1) - Divide data into training and test set
```{r}
n <- length(Stationary_series)
n

train_size <- floor(0.8 * n)

train <- Stationary_series[1:train_size]
test <- Stationary_series[(train_size + 1):n]
```
##(2) - estimate an ARMA(p,q) model with macimum likelihood
```{r}
model1 <- arima(train, order = c(3, 0, 4), method = "ML")
summary(model1) #ARMA(3.4)
```
This is an ARMA(3,4) and the aic is around 734. 

##(3) - Estimate a nested model
```{r}
model2 = arima(train, order = c(2,0,3),
      method = "ML",
      optim.control = list(maxit = 2000))
model2 #ARMA(2,3)
```
This is an ARMA(2,3), which is a nested model and it gives a lower aic (711) than the higher order model, which means it is likely optimal (lower aic should give for a better model)

##(4) - present the estimated parameters and standard errors

```{r}

coef(model1)
sqrt(diag(model1$var.coef))

coef(model2)
sqrt(diag(model2$var.coef))

```
##Create a table of estimates

```{r}
params1 <- coef(model1)
se1 <- sqrt(diag(model1$var.coef))

params2 <- coef(model2)
se2 <- sqrt(diag(model2$var.coef))

#params3 <- coef(model3)  #model 3 not found
#se3 <- sqrt(diag(model3$var.coef))

table1 <- data.frame(
  Parameter = names(params1),
  Estimate = params1,
  StdError = se1
)

table2 <- data.frame(
  Parameter = names(params2),
  Estimate = params2,
  StdError = se2
)

table1
table2
```
Although the ARMA(3,4) model converged, it is not reliable because several MA coefficients have undefined (NaN) standard errors, indicating numerical instability and issues with invertibility. Additionally, some parameter estimates are extremely large and close to non-stationary boundaries.
The ARMA(2,3) model, while still showing relatively large coefficients, provides finite standard errors for all parameters, making it more stable and interpretable.
Therefore, the ARMA(2,3) model is preferred over the ARMA(3,4) model, although both models may be overparameterized.
## Question 8
Delivery Data
```{r}
## Computing AIC
aic_11 <- AIC(fit_11)
aic_12 <- AIC(fit_12)
aic_13 <- AIC(fit_13)


## Computing BIC
bic_11 <- BIC(fit_11)
bic_12 <- BIC(fit_12)
bic_13 <- BIC(fit_13)

ic_table <- matrix(
c(aic_11, aic_12,aic_13,
bic_11, bic_12, bic_13),
nrow = 2,
byrow = TRUE)
rownames(ic_table) <- c("AIC (Train)", "BIC (Train)")
colnames(ic_table) <- c("ARMA(1,2)", "ARMA(1,1)", "MA(2)")
stargazer(ic_table,
type = "latex",
summary = FALSE,
title = "Information Criteria",
digits = 3,
header = FALSE
)

#Both the AIC and BIc tests favour the ARMA(1,2). The ARMA(1,1) and MA(2) have similar results, but the MA(2) is better than the ARMA(1,1). We will do future comparisons using only the ARMA(1,2) and MA(2).

## Likelihood Ratio Test
#H0: the constrained model (MA(2))) is adequate.
#H1: the unconstrained model (ARMA(1,2)) is significantly better.

LL13 <- fit_13$loglik  #MA(2)
LL11 <- fit_11$loglik   #ARMA(1,2)

LR_stat <- 2 * (LL11 - LL13)
df <- length(fit_11$coef) - length(fit_13$coef)
p_value <- 1 - pchisq(LR_stat, df)
lr_table <- matrix(c(LR_stat, df, p_value), nrow = 1)
colnames(lr_table) <- c("LR Statistic", "df", "p-value")
stargazer(lr_table,
type = "latex",
summary = FALSE,
title = "Likelihood Ratio Test",
digits = 4,
header = FALSE)

#The p-value is very small, so we can reject the null hypothesis at a confidence level of p=0.01 So we prefer the ARMA(1,2) over the MA(2).




#Compute the in-sample fitted values using the training set
fitted_value_11 <- y_train - residuals(fit_11)
fitted_value_13 <- y_train - residuals(fit_13)


#Produce out-of-sample forecasts on the test set 
fc_11 <- forecast(y_test, model = fit_11, h=n-n_train+1)
fc_13 <- forecast(y_test, model = fit_13, h=n-n_train+1)

#Compute the mean squared error (MSE) for both in-sample fitting and out-of-sample forecasting
rmse <- function(actual, forecast) { sqrt(mean((actual - forecast)^2)) }  #root mean squared error
mae <- function(actual, forecast) { mean(abs(actual - forecast)) }    #mean absolute error
fc_11

# Performance table
perf_table_d <- matrix(c(rmse(y_train, fitted_value_11), rmse(y_train, fitted_value_13),
                         rmse(y_test, fc_11$mean), rmse(y_test, fc_13$mean),
                         mae(y_train, fitted_value_11), mae(y_train, fitted_value_13),
                         mae(y_test, fc_11$mean), mae(y_test, fc_13$mean)
                         ), nrow = 4, byrow = TRUE)
rownames(perf_table_d) <- c("RMSE (In-sample)", "RMSE (Out-of-sample)",
"MAE (In-sample)", "MAE (Out-of-sample)")
colnames(perf_table_d) <- c("ARMA(1,2)", "MA(2)")
stargazer(perf_table_d,
type = "latex",
summary = FALSE,
title = "Performance Comparison",
digits = 4,
header = FALSE)

#For the out-of-sample results, the ARMA(1,2) has smaller errors for both the RMSE and MAE, so it is better fitted.


#Each of the tests above favoured the ARMA(1,2) model over the MA(2).So we recommend an ARMA(1,2) for this series.
```

Price Data
```{r}

#8.1
#Computing AIC and BIC training models
library(stargazer)
library(forecast)
model34=model1
model23=model2
#model 1& 2 are ARMA(3,4) and ARMA(2,3) respectively

aic_34=AIC(model34)
aic_23=AIC(model23)
bic_34=BIC(model34)
bic_23=BIC(model23)

ic2_table=matrix(
  c(aic_34,aic_23,
    bic_34,bic_23),
  nrow = 2,
  byrow = TRUE)


rownames(ic2_table)=c("AIC - Train","BIC - Train")
colnames(ic2_table)=c("ARMA(3,4)","ARMA(2,3)")

print(ic2_table)

#comparing the AIC & BIC for train model for both ARMA(3,4) & ARMA(2,3) process. Model23, with ARMA(2,3) process has lower values for AIC and BIC. Hence Model23 is preferred for the training set.

stargazer(ic2_table,
          type= "Latex",
          summary= FALSE,
          title= "Information Criteria",
          digits=3,
          header= FALSE
          )

#test models

test_model34=arima(test, order = c(3,0,4), method="ML", include.mean = FALSE)

test_model34

test_model23=arima(test, order = c(2,0,3),method="ML", include.mean = FALSE)

test_model23

aic_34test=AIC(test_model34)
aic_23test=AIC(test_model23)
bic_34test=BIC(test_model34)
bic_23test=BIC(test_model23)

ic2test_table=matrix(
  c(aic_34test,aic_23test,
    bic_34test,bic_23test),
  nrow = 2,
  byrow = TRUE)

rownames(ic2test_table)=c("AIC - Test","BIC - Test")
colnames(ic2test_table)=c("ARMA(3,4)","ARMA(2,3)")

print(ic2test_table)

stargazer(ic2test_table,
          type= "Latex",
          summary= FALSE,
          title= "Information Criteria",
          digits=3,
          header= FALSE
          )

# Test set show lower value of AIC & BIC for test_model34, hence it is preferred.

```
```{r}
#8.2 The likelihood ratio statistic for the nested models

#Here ARMA(2,3) is a special case of ARMA(3,4)
#H0: The constraint model ARMA(2,3) is adequate
#H1: The unconstrained model ARMA(3,4) is significantly better

LL34=model34$loglik#ARMA(3,4)
LL23=model23$loglik#ARMA(2,3)

LR_stat=2*(LL23-LL34)
df=length(model34$coef)-length(model23$coef)
p_value=1-pchisq(LR_stat,df)

lr_table=matrix(c(LR_stat,df,p_value), nrow = 1)
colnames(lr_table)=c("LR Statistic", "df", "p-value")
stargazer(lr_table,
         type="latex",
         summary= FALSE,
         title="Likelihood Ratio Test",
         digits=4,
          header= FALSE)

# With p-value=1 Hence, we cannot reject H0 in favor of H1
#We have significant evidence that ARMA(2,3) is an adequate model


```
```{r}
#8.3
#compute in sample fitted value using the training set
fc_34=numeric(length(test))
fc_23=numeric(length(test))

for (i in 1:length(test)){
  price_expanding=Stationary_series[1:(train_size+i-1)]
  fit1=arima(price_expanding,order = c(3,0,4),
             method="ML", include.mean = FALSE,
             optim.control = list(maxit = 1000))
  
  fit2=arima(price_expanding,order=c(2,0,3),
             method="ML", include.mean = FALSE,
             optim.control = list(maxit = 1000))
  
  fc_34[i]=predict(fit1,n.ahead=1)$pred
  fc_23[i]=predict(fit2,n.ahead=1)$pred

}



#In- sample fitted value
fitted_34=train-residuals(model34)
fitted_23=train-residuals(model23)

#out-of-sample forecast on the test set
fc_23=forecast(test, model=model23,h=n-train_size)
fc_34=forecast(test, model=model34,h=n-train_size)


rmse=function(actual,forecast){sqrt(mean((actual-forecast)^2))}
mae=function(actual,forecast){mean(abs(actual-forecast))}

#performance table
perf_table_p=matrix(c(rmse(train,fitted_23),rmse(train,fitted_34),
                    rmse(test,fc_23$mean),rmse(test,fc_34$mean),
                    mae(train,fitted_23),mae(train,fitted_34),
                    mae(test,fc_23$mean),mae(test,fc_34$mean)),
                  nrow = 4,
                  byrow = TRUE)

rownames(perf_table_p) <- c("RMSE (In-sample)",
                          "RMSE(Out-of-sample)",
                          "MAE (In-sample)", 
                          "MAE (Out-of-sample)")
colnames(perf_table_p) <- c("ARMA(2,3)", "ARMA(3,4)")
stargazer(perf_table_p,
          type = "latex",
          summary = FALSE,
          title = "Performance Comparison",
          digits = 4,
          header = FALSE)


#Both RMSE and MAE- In sample- are smaller with ARMA(2,3)
#Therefore ARMA(2,3) is preferred
#Note: oddly enough - for ARMA(3,4) Out sample error is smaller than in sample, it could be due the smaller sample size of the test set

```
Based on the diagnostic tests above, ARMA(2,3) is the better recommendation.

##Question 9

Delivery Data
```{r}
#approximate the innovation series using the innovations algorithm
# Innovations
e_hat <- residuals(fit_11)
  
# Properties of innovations
mean=mean(e_hat)
print(mean)
var=var(e_hat)
print(var)
autocorrelation = acf(e_hat)
#The mean of the innovation is close to zero, the variance is less than infinity, and the autocorrelation is not significant for lags other than 0, so it behaves as white noise.


```

Price Data
```{r}

#9.1 Approximate the innovation series
library(itsmr)
# Fit
fit <- ar(fitted_23, aic = TRUE, order.max = 30,
          method ="yule-walker")
# Innovations
e_hat <- residuals(fit)

#9.2 Statistical properties of the approximated innovation

model23_mean=mean(e_hat, na.rm = TRUE)
model23_var=var(e_hat,na.rm = TRUE)
x=acf(e_hat, lag.max = NULL, type = c("correlation", "covariance", "partial"), plot = TRUE, na.action = na.pass)



```




