---
title: "Project Pt2 Price"
author: "Group 1"
date: "2026-03-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pre-Project Part 2 filtering and processing
```{r}
library(patchwork)
library(tseries)
library(gridExtra)
        
#2.1 Removing Trend

#converting time series to object

Price_data_ts=ts(Price_data$avg_price_canada,
                 start = c(2002-01-01,1),
                 frequency = 12)

str(Price_data_ts)

#calculate transformation
Diff=diff(Price_data_ts)#First difference
Log_diff=diff(log(Price_data_ts))#Difference of Logarithms
#2.2 Removing seasonal component

library(forecast)


# Removing Seasonality

P_D1<-diff(Log_diff,difference=1)
P_D3<-diff(Log_diff,differences=3)
P_D12<-diff(Log_diff,differences=12)
P_D24<-diff(Log_diff,differences=24)


#preference to Difference of 1st order

Deseasonlised_pricedata=P_D1

# Apply filtering to log-differenced series
Series_to_filter <- Price_data_ts
n <- length(Series_to_filter)

# --- q = 6 ---
q1 <- 6 #this is in the range from 5 to 9, helping reduce the high volatility of the data
weights1 <- rep(1/(2*q1+1), 2*q1+1)

Linear_filtered_q6 <- stats::filter(
  Series_to_filter,
  filter = weights1,
  sides = 2
)

# --- q = 12 ---
q2 <- 12 #this helps to address the seasonality of the data
weights2 <- rep(1/(2*q2+1), 2*q2+1)

Linear_filtered_q12 <- stats::filter(
  Series_to_filter,
  filter = weights2,
  sides = 2
)
```

# Question 5: Dickey-Fuller Unit Root Test

```{r}
# ADF Test
adf.test(Stationary_series)
```
p-value = 0.01 < 0.05 and thus we confirm that the data is stationary with 95% certainty
# Question 6: Selection of ARMA Model Orders
##(1) - ACF and PACF and (2) - CIs
Autocorrelation Function (ACF):
$$ρ^​(h)=∑t=1n​(Xt​−Xˉ)2∑t=h+1n​(Xt​−Xˉ)(Xt−h​−Xˉ)​$$
```{r}
ACF_estimator <- function(X, hmax){
  n <- length(X)
  X_bar <- mean(X)
  acf_vals <- numeric(hmax+1)
  
  for(h in 0:hmax){
    num <- sum((X[(h+1):n]-X_bar)*
                 (X[1:(n-h)]-X_bar))
    den <- sum((X-X_bar)^2)
    acf_vals[h+1] <- num/den
  }
  
  return(acf_vals)
}

hmax <- 40
acf_values <- ACF_estimator(Stationary_series, hmax)

plot_acf_ci <- function(acf_values, series, hmax, ci = 0.95) {
  
  n <- length(series)
  
  # get correct z value
  z <- qnorm(1 - (1 - ci) / 2)
  
  # confidence interval
  CI <- z / sqrt(n)
  
  plot(0:hmax, acf_values, type = "h",
       main = paste("Estimated ACF (", ci*100, "% CI)", sep=""),
       xlab = "Lag", ylab = "ACF")
  
  abline(h = 0)
  abline(h = CI, col = "red", lty = 2)
  abline(h = -CI, col = "red", lty = 2)
}
# Plots for different CIs
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.90)
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.95)
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.99)

```
Lags for which coefficients not statistically different from 0:
90%: 0,1,2,12,22,24,26,30,36,38; 95%: 0,1,2,12,22,24,30,36,38; 99%: 1,2,12,22,24,36,38.

Partial Autocorelation function (PACF):
```{r}
PACF_estimator <- function(X, hmax){
  pacf_vals <- numeric(hmax)
  
  for(h in 1:hmax){
    model <- ar(X, order.max=h, aic=FALSE, method="yw")
    pacf_vals[h] <- model$ar[h]
  }
  
  return(pacf_vals)
}

# estimator with chosen CIs
plot_pacf_ci <- function(pacf_values, series, hmax, ci = 0.95){
  
  n <- length(series)
  z <- qnorm(1 - (1 - ci)/2)
  CI <- z / sqrt(n)
  
  plot(1:hmax, pacf_values, type="h",
       main=paste("Estimated PACF (", ci*100, "% CI)", sep=""),
       xlab="Lag", ylab="PACF")
  
  abline(h=0)
  abline(h=CI, col="blue", lty=2)
  abline(h=-CI, col="blue", lty=2)
}
pacf_values <- PACF_estimator(Stationary_series, hmax)

plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.90)
plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.95)
plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.99)

```
Lags for which coefficients not statistically different from 0:
90%:1,2,3,4,5,6,9,10,11,12,22,23,25,29,31,35,37; 95%: 1,2,3,4,9,10,11,12,22,35; 99%:1,2,3,4,10,11,35.
##(3) - Hypothesis regarding appropriate orders of the ARMA(p,q)
```{r}
auto.arima(Stationary_series, seasonal = FALSE, stationary =TRUE)
```
The test says that this series matches an MA(1) model
#Question 7
Checking for structural breaks
```{r}
plot(Stationary_series, type="l")
```
##(1) - Divide data into training and test set
```{r}
n <- length(Stationary_series)

train_size <- floor(0.8 * n)

train <- Stationary_series[1:train_size]
test <- Stationary_series[(train_size + 1):n]
```
##(2) - estimate an ARMA(p,q) model with macimum likelihood
```{r}
model1 <- arima(train, order = c(3, 0, 2), method = "ML")
summary(model1)
```
This is an ARMA(3,2)
##(3) - Estimate a nested model
```{r}
model2 <- arima(train, order = c(0,0,1), method="ML")
summary(model2)
```
This is an MA(1)
##(4) - present the estimated parameters and standard errors

```{r}
coef(model1)
sqrt(diag(model1$var.coef))

coef(model2)
sqrt(diag(model2$var.coef))
```
##Create a table of estimates

```{r}
params1 <- coef(model1)
se1 <- sqrt(diag(model1$var.coef))

params2 <- coef(model2)
se2 <- sqrt(diag(model2$var.coef))

table1 <- data.frame(
  Parameter = names(params1),
  Estimate = params1,
  StdError = se1
)

table2 <- data.frame(
  Parameter = names(params2),
  Estimate = params2,
  StdError = se2
)

table1
table2

```
All parameters appear statistically significant for all models since ∣estimate∣>2×SE
