---
title: "Project Pt2 Price"
author: "Group 1"
date: "2026-03-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pre-Project Part 2 filtering and processing
```{r}
library(patchwork)
library(tseries)
library(gridExtra)
library(readr)
library(ggplot2)
library(quantmod) # loading data sets from online
library(dplyr) # for the pipes
library(kableExtra) # for kable
library(lubridate) # for years...
library(knitr)
library(tidyr)

Data_Price <- read.csv("https://raw.githubusercontent.com/stardustsunrise/Time_Series/refs/heads/main/3210007701_databaseLoadingData.csv")

head(Data_Price)
str(Data_Price)

print(Data_Price)


#Convert to data frame


Price_data = data.frame(Date= as.Date(paste0(Data_Price$REF_DATE,"-01")), 
                        Price_per_MTonne= Data_Price$VALUE,
                        
                        Product=c("Wheat (except durum wheat) [1121111]"))
print(Price_data)

# Averaging the price of "Wheat (except durum wheat) [1121111]" accross all provinces to have 1 avg monthly price value for whole of Canada
Price_data<-Price_data%>%
  group_by(Date)%>%
  summarise(avg_price_canada=mean(Price_per_MTonne,na.rm= TRUE))%>%
  mutate(Months=row_number(),
         Year=year(Date),
         Quarter=quarter(Date))%>%
  ungroup()

print(Price_data)
        
#2.1 Removing Trend

#converting time series to object

Price_data_ts=ts(Price_data$avg_price_canada,
                 start = c(2002-01-01,1),
                 frequency = 12)

str(Price_data_ts)

#calculate transformation
Diff=diff(Price_data_ts)#First difference
Log_diff=diff(log(Price_data_ts))#Difference of Logarithms
#2.2 Removing seasonal component

library(forecast)


# Removing Seasonality

P_D1<-diff(Log_diff,difference=1)
P_D3<-diff(Log_diff,differences=3)
P_D12<-diff(Log_diff,differences=12)
P_D24<-diff(Log_diff,differences=24)


#preference to Difference of 12th order

Deseasonlised_pricedata=P_D12

# Apply filtering to log-differenced series
Series_to_filter <- Price_data_ts
n <- length(Series_to_filter)

# --- q = 6 ---
q1 <- 6 #this is in the range from 5 to 9, helping reduce the high volatility of the data
weights1 <- rep(1/(2*q1+1), 2*q1+1)

Linear_filtered_q6 <- stats::filter(
  Series_to_filter,
  filter = weights1,
  sides = 2
)

# --- q = 12 ---
q2 <- 12 #this helps to address the seasonality of the data
weights2 <- rep(1/(2*q2+1), 2*q2+1)

Linear_filtered_q12 <- stats::filter(
  Series_to_filter,
  filter = weights2,
  sides = 2
)
```
```{r}
#compare the differences
plot(P_D1)
plot(P_D12)
```
```{r}
library(forecast)
library(gridExtra)
library(ggplot2)

# Function to simulate and return ACF/PACF plots
plot_arma_acf_pacf <- function(ar = numeric(0), ma = numeric(0), n = 500) {
  # Use safe coefficients if ar/ma supplied are too big
  if(length(ar) > 0) ar <- rep(0.3, length(ar))
  if(length(ma) > 0) ma <- rep(0.5, length(ma))
  
  ts_data <- arima.sim(model = list(ar = ar, ma = ma), n = n)
  acf_plot <- ggAcf(ts_data) + ggtitle(paste0("ACF AR=", length(ar), " MA=", length(ma)))
  pacf_plot <- ggPacf(ts_data) + ggtitle(paste0("PACF AR=", length(ar), " MA=", length(ma)))
  return(list(acf_plot, pacf_plot))
}

# Example: simulate ARMA(2,1)
plots <- plot_arma_acf_pacf(ar = c(0.3, 0.3), ma = 0.5)
grid.arrange(plots[[1]], plots[[2]], ncol = 2)
```


# Question 5: Dickey-Fuller Unit Root Test

```{r}

Stationary_series <- Deseasonlised_pricedata
library(tseries)
# ADF Test
adf.test(Stationary_series)
```
p-value = 0.01 < 0.05 and thus we confirm that the data is stationary with 95% certainty
# Question 6: Selection of ARMA Model Orders
##(1) - ACF and PACF and (2) - CIs
Autocorrelation Function (ACF):
$$ρ^​(h)=∑t=1n​(Xt​−Xˉ)2∑t=h+1n​(Xt​−Xˉ)(Xt−h​−Xˉ)​$$
```{r}
ACF_estimator <- function(X, hmax){
  n <- length(X)
  X_bar <- mean(X)
  acf_vals <- numeric(hmax+1)
  
  for(h in 0:hmax){
    num <- sum((X[(h+1):n]-X_bar)*
                 (X[1:(n-h)]-X_bar))
    den <- sum((X-X_bar)^2)
    acf_vals[h+1] <- num/den
  }
  
  return(acf_vals)
}

hmax <- 40
acf_values <- ACF_estimator(Stationary_series, hmax)

plot_acf_ci <- function(acf_values, series, hmax, ci = 0.95) {
  
  n <- length(series)
  
  # get correct z value
  z <- qnorm(1 - (1 - ci) / 2)
  
  # confidence interval
  CI <- z / sqrt(n)
  
  plot(0:hmax, acf_values, type = "h",
       main = paste("Estimated ACF (", ci*100, "% CI)", sep=""),
       xlab = "Lag", ylab = "ACF")
  
  abline(h = 0)
  abline(h = CI, col = "red", lty = 2)
  abline(h = -CI, col = "red", lty = 2)
}
# Plots for different CIs
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.90)
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.95)
plot_acf_ci(acf_values, Stationary_series, hmax, ci = 0.99)

```
Lags for which coefficients not statistically different from 0:
90%: 0,1,2,12,22,24,26,30,36,38; 95%: 0,1,2,12,22,24,30,36,38; 99%: 1,2,12,22,24,36,38.

Partial Autocorelation function (PACF):
```{r}
PACF_estimator <- function(X, hmax){
  pacf_vals <- numeric(hmax)
  
  for(h in 1:hmax){
    model <- ar(X, order.max=h, aic=FALSE, method="yw")
    pacf_vals[h] <- model$ar[h]
  }
  
  return(pacf_vals)
}

# estimator with chosen CIs
plot_pacf_ci <- function(pacf_values, series, hmax, ci = 0.95){
  
  n <- length(series)
  z <- qnorm(1 - (1 - ci)/2)
  CI <- z / sqrt(n)
  
  plot(1:hmax, pacf_values, type="h",
       main=paste("Estimated PACF (", ci*100, "% CI)", sep=""),
       xlab="Lag", ylab="PACF")
  
  abline(h=0)
  abline(h=CI, col="blue", lty=2)
  abline(h=-CI, col="blue", lty=2)
}
pacf_values <- PACF_estimator(Stationary_series, hmax)

plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.90)
plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.95)
plot_pacf_ci(pacf_values, Stationary_series, hmax, ci=0.99)

```
Lags for which coefficients not statistically different from 0:
90%:1,2,3,4,5,6,9,10,11,12,22,23,25,29,31,35,37; 95%: 1,2,3,4,9,10,11,12,22,35; 99%:1,2,3,4,10,11,35.


##(3) - Hypothesis regarding appropriate orders of the ARMA(p,q)
```{r}

auto.arima(Stationary_series, seasonal = FALSE, stationary =TRUE)
```
The test says that this series matches an AR(5) model. However,the graphs suggest more of a ARMA(3,4) model (4 significant lags on ACF, 3 significant lags on PACF). We will try to go off of the graph for the sake of this analysis. 
#Question 7
Checking for structural breaks
```{r}
plot(Stationary_series, type="l")
```
While there are some spikes in the data, it does not appear that there were any structural breaks. Thus, we can proceed with an 80-20 division for the data
##(1) - Divide data into training and test set
```{r}
n <- length(Stationary_series)
n

train_size <- floor(0.8 * n)

train <- Stationary_series[1:train_size]
test <- Stationary_series[(train_size + 1):n]
```
##(2) - estimate an ARMA(p,q) model with macimum likelihood
```{r}
model1 <- arima(train, order = c(3, 0, 4), method = "ML")
summary(model1) #ARMA(3.4)
```
This is an ARMA(3,4) and the aic is around 734. 

##(3) - Estimate a nested model
```{r}
model2 = arima(train, order = c(2,0,3),
      method = "ML",
      optim.control = list(maxit = 2000))
model2 #ARMA(2,3)
```
This is an ARMA(2,3), which is a nested model and it gives a lower aic (711) than the higher order model, which means it is likely optimal (lower aic should give for a better model)

##(4) - present the estimated parameters and standard errors

```{r}

coef(model1)
sqrt(diag(model1$var.coef))

coef(model2)
sqrt(diag(model2$var.coef))

```
##Create a table of estimates

```{r}
params1 <- coef(model1)
se1 <- sqrt(diag(model1$var.coef))

params2 <- coef(model2)
se2 <- sqrt(diag(model2$var.coef))

params3 <- coef(model3)  #model 3 not found
se3 <- sqrt(diag(model3$var.coef))

table1 <- data.frame(
  Parameter = names(params1),
  Estimate = params1,
  StdError = se1
)

table2 <- data.frame(
  Parameter = names(params2),
  Estimate = params2,
  StdError = se2
)

table1
table2
```
Although the ARMA(3,4) model converged, it is not reliable because several MA coefficients have undefined (NaN) standard errors, indicating numerical instability and issues with invertibility. Additionally, some parameter estimates are extremely large and close to non-stationary boundaries.
The ARMA(2,3) model, while still showing relatively large coefficients, provides finite standard errors for all parameters, making it more stable and interpretable.
Therefore, the ARMA(2,3) model is preferred over the ARMA(3,4) model, although both models may be overparameterized.

```{r}

#8.1
#Computing AIC and BIC training models
library(stargazer)
library(forecast)
aic_21=AIC(model1)
aic_22=AIC(model2)
bic_21=BIC(model1)
bic_22=BIC(model2)

ic2_table=matrix(
  c(aic_21,aic_22,
    bic_21,bic_22),
  nrow = 2,
  byrow = TRUE)

rownames(ic2_table)=c("AIC - Train","BIC - Train")
colnames(ic2_table)=c("ARMA(3,4)","ARMA(2,3)")

print(ic2_table)

stargazer(ic2_table,
          type= "Latex",
          summary= FALSE,
          title= "Information Criteria",
          digits=3,
          header= FALSE
          )

test_model21=arima(test, order = c(3,0,4), include.mean = FALSE)
test_model22=arima(test, order = c(2,0,3), include.mean = FALSE)
test_model22 <- arima(test, order = c(3, 0, 4), method = "ML")
test_model22= arima(test, order = c(2,0,3),
      method = "ML",
      optim.control = list(maxit = 2000))


```
```{r}
#8.2 The likelihood ratio statistic for the nested models

#Here ARMA(2,3) is a special case of ARMA(3,4)
#H0: The constraint model ARMA(2,3)
#H1: The unconstraint model ARMA(3,4)

LL21=model1$loglik#ARMA(3,4)
LL22=model2$loglik#ARMA(2,3)

LR_stat=2*(LL21-LL22)
df=length(model2$coef)-length(model2$coef)
p_value=1-pchisq(LR_stat,df)

lr_table=matrix(c(LR_stat,df,p_value), nrow = 1)
colnames(lr_table)=c("LR Statistic", "df", "p-value")
stargazer(lr_table,
         type="latex",
         summary= FALSE,
         title="Likelihood Ratio Test",
         digits=4,
          header= FALSE)

# With p-value=1 we must reject H0. There is sufficient evidence that ARMA(2,3) is a better model


```
```{r}
#8.3
#compute in sample fitted value using the training set

#In- sample fitted value
fitted_34=train-residuals(model1)
fitted_23=train-residuals(model2)

#out-of-sample forecaste on the test set
fc_23=forecast(test, model=model2,h=n-train_size)
fc_34=forecast(test, model=model1,h=n-train_size)
fc_23=numeric(length(test))
fc_34=numeric(length(test))

rmse=function(actual,forecast){sqrt(mean((actual-forecast)^2))}
mae=function(actual,forecast){mean(abs(actual-forecast))}

#performance table
perf_table=matrix(c(rmse(train,model2),rmse(train,model1),
                    rmse(test,fc_23),rmse(test,fc_34),
                    mae(train,model2),mae(train,fc_34),
                    mae(test,fc_23),mae(test,fc_34)),nrow = 4,byrow = TRUE)

#
```




