---
title: "Time Series Project Part II (Delivery)"
author: "Group 1"
output:
  pdf_document:
    latex_engine: xelatex
date: "2026-03-14"
---

# 

```{r}

library(readr)
library(ggplot2)
library(readr)
library(ggplot2)
library(quantmod) # loading data sets from online
library(dplyr) # for the pipes
library(kableExtra) # for kable
library(lubridate) # for years...
library(tidyr)
library(knitr)
library(forecast)     # for Acf, Pacf, auto.arima, etc.  
library(tseries)      # for adf.test  
library(gridExtra)    # for arranging plots  


data2 <- read_csv("3210035101_databaseLoadingData_2.csv")
head(data2)



delivery_data = data.frame(Date = data2$REF_DATE,
                           Value = as.numeric(data2$VALUE))

head(delivery_data)
p2 = ggplot(delivery_data, aes(x=Date, y = Value, group=1))+
             geom_line(color="blue", size=1)+
  labs(
    title = "Producer deliveries of Wheat, Excluding Durum",
    subtitle = "2002-2025",
    x = "Date (Monthly)",
    y = "Grain (Tonnes)",
    caption = "Source: Statistics Canada. Table 32-10-0351-01  Producer deliveries of major grains"
  ) 
print(p2)

delivery_ts <- ts(data2$VALUE,
                  start = c(2002,1), # Start year and month: January 2002
                  frequency = 12) # 12 observations per year (monthly data)


# Define moving average filter function
moving_average <- function(X, q) {
  N <- length(X)              # Length of input series
  weight <- 1 / (2 * q + 1)   # Equal weights for symmetric MA
  indices <- (q+1):(N-q)      # Indices where filter can be applied
  X_filt <- rep(NA, N)        # Initialize output with NAs
  
  # Apply moving average filter
  for (i in indices) {
    X_filt[i] <- sum(X[(i-q):(i+q)] * weight)
  }
  return(X_filt)               # Return filtered series
}


delivery_data_D3 = diff(log(delivery_data$Value))

delivery_data_des = diff(delivery_data_D3, lag=12)

```

# Part II

## Question 5

```{r}


library(tseries)

# ADF Test
adf.test(delivery_data_des)

```

## Question 6

Point 1 and 2

```{r}
#---ACF estimator function---------------

ACF_estimator <- function(X, hmax){
      n <- length(X)
      X_bar <- mean(X)
      acf_vals <- numeric(hmax+1)
      
      for(h in 0:hmax){
        num <- sum((X[(h+1):n]-X_bar)*
                     (X[1:(n-h)]-X_bar))
        den <- sum((X-X_bar)^2)
        acf_vals[h+1] <- num/den
      }
      
      return(acf_vals)
    }


res <- na.omit(delivery_data_des)

hmax <- 40


acf_values <- ACF_estimator(res, hmax)

#---PACF estimator function-----------------------------------

PACF_estimator <- function(X, hmax){
      
      X <- na.omit(X)        # ← critical line
      pacf_vals <- numeric(hmax)
      
      for(h in 1:hmax){
        model <- ar(X, order.max = h, aic = FALSE, method = "yw")
        pacf_vals[h] <- model$ar[h]
      }
      
      return(pacf_vals)
    }


pacf_values <- PACF_estimator(delivery_data_des, hmax)

#--------Setting Up the CI--------------

CI_95 <- 1.96 / sqrt(length(res))  #90% 1.645    99% 2.576
CI_90 <- 1.645 / sqrt(length(res))
CI_99 <- 2.576 / sqrt(length(res))

#----------Plot ACF-----------------------------------

#90% CI Plot

plot(0:hmax, acf_values, type = "h",
      main = "Estimated ACF 90% CI",
      xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_90, col = "red", lty = 2)
abline(h = -CI_90, col = "red", lty = 2)


    #95% CI Plot
plot(0:hmax, acf_values, type = "h",
     main = "Estimated ACF 95% CI",
     xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_95, col = "red", lty = 2)
abline(h = -CI_95, col = "red", lty = 2)

#99% CI Plot
plot(0:hmax, acf_values, type = "h",
      main = "Estimated ACF 99% CI",
      xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_99, col = "red", lty = 2)
abline(h = -CI_99, col = "red", lty = 2)

#----------Plot PACF-------------------------

#90% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 90% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_90, col="blue", lty=2)
abline(h=-CI_90, col="blue", lty=2)


#95% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 95% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_95, col="blue", lty=2)
abline(h=-CI_95, col="blue", lty=2)


#99% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 99% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_99, col="blue", lty=2)
abline(h=-CI_99, col="blue", lty=2)




```

Point 3

The ACF shows a significant spike at Lag 1 and a potentially significant spike at Lag 2 before cutting off. This is characteristic of a Moving Average process, specifically suggesting an MA(1) or MA(2). The PACF does not show a clear cutoff; instead, we can see a decaying pattern (ignoring seasonal lags at 12, 24, and 36). This decay in the PACF is a sign of an AR(1) component or that this is simply and MA(q) process.

Significant spikes remain at seasonal lags (12, 24, 36). While these indicate remaining seasonal correlation, they are being bypassed for the scope of this project, as advised by professor Boire.

For Question 7, we compare three model ARMA(1,2), ARMA(1,1) and MA(2).

## Question 7

```{r stargazer, results='asis'}

#-----Splitting into training test and test sets----------------
n = length(delivery_data_des)

y = delivery_data_des

n_train <- floor(0.8 * n)
y_train <- y[1:n_train]
y_test <- y[(n_train+1):n]


#----Model Estimation on the Training Set--------------------

fit_11 <- arima(y_train, order = c(1,0,2), include.mean = FALSE) #fit given by 
fit_12 <- arima(y_train, order = c(1,0,1), include.mean = FALSE) #testing this fit to see if we're overfitting
fit_13 <- arima(y_train, order = c(0,0,2), include.mean = FALSE) #random test


#----Displaying the results with stargazer---------

se_11 <- sqrt(diag(fit_11$var.coef))
se_12 <- sqrt(diag(fit_12$var.coef))
se_13 <- sqrt(diag(fit_13$var.coef))

library(stargazer)

stargazer(fit_11, fit_12, fit_13,
type = "latex",
se = list(se_11, se_12, se_13),
title = "ARMA Estimation Results",
column.labels = c("ARMA(1,2)", "ARMA(1,1)", "MA(2)"),
digits = 3,
header = FALSE)




```

For the model ARMA(1,2), all of our coefficients are significative. A strong negative autoregressive coefficient AR(1)= -0.886 suggests the series is oscillating. If the value was high last month, it’s very likely to be low this month.MA(1) = 0.304 is a momentum effect and if we had a positive shock last month, approximately 30.4% of that shock persists into the next month in the same direction. The second lag MA(2) = 0.696 is the most dominant feature of the moving average component. A positive shock occurring at two months ago would results in a significant downward pressure (around 69.9% of that positive shock) on the series this month.

For the second model ARMA(1,1), The AR(1)=0.056 is not significant. This model is "over-specified." It is trying to use an autoregressive component that isn't actually helping.

For the third model, the MA(2)= 0.048 term is not significant for similar reasons as the ARMA(1,1)

The best model so far is the ARMA(1,2).

## Question 8

```{r}
## Computing AIC
aic_11 <- AIC(fit_11)
aic_12 <- AIC(fit_12)
aic_13 <- AIC(fit_13)


## Computing BIC
bic_11 <- BIC(fit_11)
bic_12 <- BIC(fit_12)
bic_13 <- BIC(fit_13)

ic_table <- matrix(
c(aic_11, aic_12,aic_13,
bic_11, bic_12, bic_13),
nrow = 2,
byrow = TRUE)
rownames(ic_table) <- c("AIC (Train)", "BIC (Train)")
colnames(ic_table) <- c("ARMA(1,2)", "ARMA(1,1)", "MA(2)")
stargazer(ic_table,
type = "latex",
summary = FALSE,
title = "Information Criteria",
digits = 3,
header = FALSE
)

#Both the AIC and BIc tests favour the ARMA(1,2). The ARMA(1,1) and MA(2) have similar results, but the MA(2) is better than the ARMA(1,1). We will do future comparisons using only the ARMA(1,2) and MA(2).

## Likelihood Ratio Test
#H0: the constrained model (MA(2))) is adequate.
#H1: the unconstrained model (ARMA(1,2)) is significantly better.

LL13 <- fit_13$loglik  #MA(2)
LL11 <- fit_11$loglik   #ARMA(1,2)

LR_stat <- 2 * (LL11 - LL13)
df <- length(fit_11$coef) - length(fit_13$coef)
p_value <- 1 - pchisq(LR_stat, df)
lr_table <- matrix(c(LR_stat, df, p_value), nrow = 1)
colnames(lr_table) <- c("LR Statistic", "df", "p-value")
stargazer(lr_table,
type = "latex",
summary = FALSE,
title = "Likelihood Ratio Test",
digits = 4,
header = FALSE)

#The p-value is very small, so we can reject the null hypothesis at a confidence level of p=0.01 So we prefer the ARMA(1,2) over the MA(2).




#Compute the in-sample fitted values using the training set
fitted_value_11 <- y_train - residuals(fit_11)
fitted_value_13 <- y_train - residuals(fit_13)


#Produce out-of-sample forecasts on the test set 
fc_11 <- forecast(y_test, model = fit_11, h=n-n_train)
fc_13 <- forecast(y_test, model = fit_13, h=n-n_train)

#Compute the mean squared error (MSE) for both in-sample fitting and out-of-sample forecasting
rmse <- function(actual, forecast) { sqrt(mean((actual - forecast)^2)) }  #root mean squared error
mae <- function(actual, forecast) { mean(abs(actual - forecast)) }    #mean absolute error


# Performance table
perf_table <- matrix(c(
rmse(y_train, fitted_value_11), rmse(y_train, fitted_value_13),
rmse(y_test, fc_11$mean), rmse(y_test, fc_13$mean),
mae(y_train, fitted_value_11), mae(y_train, fitted_value_13),
mae(y_test, fc_11$mean), mae(y_test, fc_13$mean)
), nrow = 4, byrow = TRUE)
rownames(perf_table) <- c("RMSE (In-sample)", "RMSE (Out-of-sample)",
"MAE (In-sample)", "MAE (Out-of-sample)")
colnames(perf_table) <- c("ARMA(1,2)", "MA(2)")
stargazer(perf_table,
type = "latex",
summary = FALSE,
title = "Performance Comparison",
digits = 4,
header = FALSE)

#For the out-of-sample results, the ARMA(1,2) has smaller errors for both the RMSE and MAE, so it is better fitted.


#Each of the tests above favoured the ARMA(1,2) model over the MA(2).So we recommend an ARMA(1,2) for this series.
```
##Question 9
```{r}
#approximate the innovation series using the innovations algorithm
# Innovations
e_hat <- residuals(fit_11)
  
# Properties of innovations
mean=mean(e_hat)
print(mean)
var=var(e_hat)
print(var)
autocorrelation = acf(e_hat)
#The mean of the innovation is close to zero, the variance is less than infinity, and the autocorrelation is not significant for lags other than 0, so it behaves as white noise.


```

