---
title: "Time Series Project Part II (Delivery)"
author: "Group 1"
output:
  pdf_document:
    latex_engine: xelatex
date: "2026-03-14"
---

#

```{r}

library(readr)
library(ggplot2)
library(readr)
library(ggplot2)
library(quantmod) # loading data sets from online
library(dplyr) # for the pipes
library(kableExtra) # for kable
library(lubridate) # for years...
library(tidyr)
library(knitr)
library(forecast)     # for Acf, Pacf, auto.arima, etc.  
library(tseries)      # for adf.test  
library(gridExtra)    # for arranging plots  


data2 <- read_csv("3210035101_databaseLoadingData_2.csv")
head(data2)



delivery_data = data.frame(Date = data2$REF_DATE,
                           Value = as.numeric(data2$VALUE))

head(delivery_data)
p2 = ggplot(delivery_data, aes(x=Date, y = Value, group=1))+
             geom_line(color="blue", size=1)+
  labs(
    title = "Producer deliveries of Wheat, Excluding Durum",
    subtitle = "2002-2025",
    x = "Date (Monthly)",
    y = "Grain (Tonnes)",
    caption = "Source: Statistics Canada. Table 32-10-0351-01  Producer deliveries of major grains"
  ) 
print(p2)

delivery_ts <- ts(data2$VALUE,
                  start = c(2002,1), # Start year and month: January 2002
                  frequency = 12) # 12 observations per year (monthly data)


# Define moving average filter function
moving_average <- function(X, q) {
  N <- length(X)              # Length of input series
  weight <- 1 / (2 * q + 1)   # Equal weights for symmetric MA
  indices <- (q+1):(N-q)      # Indices where filter can be applied
  X_filt <- rep(NA, N)        # Initialize output with NAs
  
  # Apply moving average filter
  for (i in indices) {
    X_filt[i] <- sum(X[(i-q):(i+q)] * weight)
  }
  return(X_filt)               # Return filtered series
}


delivery_data_D3 = diff(log(delivery_data$Value))

delivery_data_des = diff(delivery_data_D3, lag=12)

```

# Part II

## Question 5 

```{r}


library(tseries)

# ADF Test
adf.test(delivery_data_des)

```






## Question 6

Point 1 and 2


```{r}
#---ACF estimator function---------------

ACF_estimator <- function(X, hmax){
      n <- length(X)
      X_bar <- mean(X)
      acf_vals <- numeric(hmax+1)
      
      for(h in 0:hmax){
        num <- sum((X[(h+1):n]-X_bar)*
                     (X[1:(n-h)]-X_bar))
        den <- sum((X-X_bar)^2)
        acf_vals[h+1] <- num/den
      }
      
      return(acf_vals)
    }


res <- na.omit(delivery_data_des)

hmax <- 40


acf_values <- ACF_estimator(res, hmax)

#---PACF estimator function-----------------------------------

PACF_estimator <- function(X, hmax){
      
      X <- na.omit(X)        # ← critical line
      pacf_vals <- numeric(hmax)
      
      for(h in 1:hmax){
        model <- ar(X, order.max = h, aic = FALSE, method = "yw")
        pacf_vals[h] <- model$ar[h]
      }
      
      return(pacf_vals)
    }


pacf_values <- PACF_estimator(delivery_data_des, hmax)

#--------Setting Up the CI--------------

CI_95 <- 1.96 / sqrt(length(res))  #90% 1.645    99% 2.576
CI_90 <- 1.645 / sqrt(length(res))
CI_99 <- 2.576 / sqrt(length(res))

#----------Plot ACF-----------------------------------

#90% CI Plot

plot(0:hmax, acf_values, type = "h",
      main = "Estimated ACF 90% CI",
      xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_90, col = "red", lty = 2)
abline(h = -CI_90, col = "red", lty = 2)


    #95% CI Plot
plot(0:hmax, acf_values, type = "h",
     main = "Estimated ACF 95% CI",
     xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_95, col = "red", lty = 2)
abline(h = -CI_95, col = "red", lty = 2)

#99% CI Plot
plot(0:hmax, acf_values, type = "h",
      main = "Estimated ACF 99% CI",
      xlab = "Lag", ylab = "ACF")
abline(h = 0)
abline(h = CI_99, col = "red", lty = 2)
abline(h = -CI_99, col = "red", lty = 2)

#----------Plot PACF-------------------------

#90% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 90% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_90, col="blue", lty=2)
abline(h=-CI_90, col="blue", lty=2)


#95% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 95% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_95, col="blue", lty=2)
abline(h=-CI_95, col="blue", lty=2)


#99% CI Plot
plot(1:hmax, pacf_values, type="h",
      main="Estimated PACF 99% CI",
      xlab="Lag", ylab="PACF")
abline(h=0)
abline(h=CI_99, col="blue", lty=2)
abline(h=-CI_99, col="blue", lty=2)




```



Point 3


The ACF shows a significant spike at Lag 1 and a potentially significant spike at Lag 2 before cutting off. This is characteristic of a Moving Average process, specifically suggesting an MA(1) or MA(2). The PACF does not show a clear cutoff; instead, we can see a  decaying pattern (ignoring seasonal lags at 12, 24, and 36). This decay in the PACF is a sign of an AR(1) component or that this is simply and MA(q) process.

Significant spikes remain at seasonal lags (12, 24, 36). While these indicate remaining seasonal correlation, they are being bypassed for the scope of this project, as advised by professor Boire.

For Question 7, we compare three model ARMA(1,2), ARMA(1,1) and MA(2). 


## Question 7

```{r stargazer, results='asis'}

#-----Splitting into training test and test sets----------------
n = length(delivery_data_des)

y = delivery_data_des

n_train <- floor(0.8 * n)
y_train <- y[1:n_train]
y_test <- y[(n_train+1):n]


#----Model Estimation on the Training Set--------------------

fit_11 <- arima(y_train, order = c(1,0,4), include.mean = FALSE) #fit given by 
fit_12 <- arima(y_train, order = c(1,0,2), include.mean = FALSE) #testing this fit to see if we're overfitting
fit_13 <- arima(y_train, order = c(0,0,2), include.mean = FALSE) #random test


#----Displaying the results with stargazer---------

se_11 <- sqrt(diag(fit_11$var.coef))
se_12 <- sqrt(diag(fit_12$var.coef))
se_13 <- sqrt(diag(fit_13$var.coef))

library(stargazer)

stargazer(fit_11, fit_12, fit_13,
type = "latex",
se = list(se_11, se_12, se_13),
title = "ARMA Estimation Results",
column.labels = c("ARMA(1,4)", "ARMA(1,1)", "MA(2)"),
digits = 3,
header = FALSE)




```


For the model ARMA(2,1), none of the coefficients are significant. 

For the second model ARMA(1,1), the coefficients are all significant which might indicate a good fit. The AR(1)= 0.263 indicates a positive momentum. If the series was high in the previous period, it has a 26.3% tendency to stay high in the next period. The MA(1) = -0.926 represents a strong shock absorbed. This suggests that if there was an error spike in the previous period, the model expects an almost equal and opposite correction in the next period, which we can sometimes see in the raw data. 

For the third model MA(2), we also have all significant coefficients. MA(1)= -0.648 can be interpreted that if there was a positive shock last month, this month is adjusted downwards by 64.8% of last months value. MA(2)=-0.246 is interpreted as the secondary correction based on the shock two months prior. 

